Using Incremental Many-to-One Queries to Build a Fast and Tight Heuristic for A* in Road Networks

### 2022-07-21 paper/ch_potentials.tex
update affiliation
relaxation fix
integrate a lot of reviewer feedback

\newcommand*{\dist}{\mathcal{D}}

@@

\author{Ben Strasser}
\email{academia@ben-strasser.net}
\affiliation{}

@@

\begin{abstract}
We study exact, efficient and practical algorithms for route planning applications in large road networks.
Routing applications often require integrating the current traffic situation, planning ahead with traffic predictions for the future, respecting forbidden turns, and many other features depending on the exact application.
While Dijkstra's algorithm can be used to solve these problems, it is too slow for many applications.
A* is a classical approach to accelerate Dijkstra's algorithm.
A* can support many extended scenarios without much additional implementation complexity.
However, A*'s performance depends on the availability of a good heuristic that estimates distances.
Computing tight distance estimates is a challenge on its own.
On road networks, shortest paths can also be quickly computed using hierarchical speedup techniques.
They achieve speed and exactness but sacrifice A*'s flexibility.
Extending them to certain practical applications can be hard.
In this paper, we present Lazy RPHAST, an algorithm to efficiently compute distances from many nodes towards a single target and use it as an A* heuristic.
As it uses Contraction Hierarchies (CH), a hierarchical technique, we call our A* heuristic CH-Potentials.
Our approach allows decoupling the supported applications from the hierarchical speed-up technique.
Additionally, we describe A* optimizations to accelerate the processing of low degree nodes, which often occur in road networks and present a new pruning criterion for symmetrical bidirectional A*.
An extensive experimental study confirms the practicality of our approach for many applications.
\end{abstract}

@@

\section{Introduction}
\label{sec:intro}
The past decade has seen a plethora of research on route planning in large street networks~\cite{bdgmpsww-rptn-16}.
Routing a user through a road network can be formalized as the shortest path problem in weighted graphs.
Nodes represent intersections.
Roads are modeled using edges.
Edges are weighted by their traversal times.
The problem can be solved with Dijkstra's algorithm~\cite{d-ntpcg-59}.
Unfortunately, on continental sized networks, it is too slow for many applications.
Thus, speed-up techniques have been developed.
One popular example are Contraction Hierarchies~(CH)~\cite{gssv-erlrn-12}.
They have been used successfully in many real world applications.
A CH exploits the inherent hierarchy of road networks.
In a preprocessing step, additional shortcut edges are inserted, which allow skipping unimportant parts of the network at query time.
Another popular example is Multi-Level-Dijkstra~(MLD)~\cite{swz-umlgt-02} also known as CRP~\cite{dgpw-crprn-13}.
It is also used in practice~\cite{bingblog}.
MLD also uses shortcut edges.
Both approaches achieve speed-ups of at least three orders of magnitude over Dijkstra's algorithm.

\begin{figure}
\centering
\includegraphics[width=.6\columnwidth]{fig/searchspace_st.png}
\caption{Nodes explored by A*. Color indicates the node removal order from the queue. Blue was removed first. Next is green. Red was removed last.}
\label{img:search-space}
\Description[A map where nodes visited by A* are marked.]{
A map of Karlsruhe where nodes visited by A* are marked.
The shape of the area with the visited nodes is similar to a drop.
The source node roughly lies in the center of the drop, the target node at the tip.
}
\end{figure}

Unfortunately, for many real world applications, this basic graph model is too simplistic.
For realistic routing, many additional features need to be considered.
This includes turn costs and restrictions, live traffic, user preferences, and traffic predictions. %, temporary driving bans for certain types of vehicles.
Some applications may have additional application-specific requirements.
Extending Dijkstra's algorithm to support these features is usually easy.
Extending hierarchical speed-up techniques is often possible, too.
However, the algorithm development is vastly more complex.
For every feature, dedicated research papers exist that extend CH and other approaches.
Supporting the combination of several features is even harder.
For example, we are not aware of any work combining all features mentioned above.
In this article, we describe an algorithmic building block, that allows handling the combination of all above-mentioned features -- and probably more.

Our approach decouples the development of extensions from the hierarchical speed-up technique by utilizing the A* algorithm~\cite{hnr-afbhd-68}.
A* is a goal-directed variant of Dijkstra's algorithm.
See Figure~\ref{img:search-space} for an example of nodes traversed during an A* search.
A* uses a \emph{heuristic} to guide the search towards the goal.
A heuristic is a function that maps a node to an estimate of the remaining distance to the goal.
A*'s running time crucially depends on how tight this estimate is.
Further, evaluating the heuristic must be fast.
% The free flow travel time to the goal is a valid heuristic for most features.
% In many settings, it is the tightest lower bound available during preprocessing, i.e., independent of real-time data and user settings.
% However, to be useful we must be able to evaluate the free flow heuristic \todo{ist free flow heuristic besser als tight heuristic?} quickly and exactly.
In this article, we introduce Lazy RPHAST and build on it CH-Potentials, a fast heuristic with tight estimates.
Internally, the heuristic uses a CH.
Fortunately, this is an implementation detail from the perspective of the A*.
To support a new feature, we only need to modify the A* algorithm.
The heuristic core containing the CH remains untouched.
Extending A* is vastly easier than extending a CH.
This enables us to design algorithms for a multitude of features.

This article is the extended version of our conference paper ``A Fast and Tight Heuristic for A* in Road Networks''~\cite{strasser_et_al:LIPIcs.SEA.2021.6}.
The paper introduced the basic CH-Potentials heuristic, some query optimizations for handling of low-degree nodes, common in road networks, and some applications of CH-Potentials.
In this article, we discuss and evaluate Lazy RPHAST, the central ingredient of CH-Potentials, in greater detail and introduce an improved variant for Customizable Contraction Hierarchies.
% Furthermore, we evaluate Lazy RPHAST in the context of one-to-many and many-to-many problems.
The original paper only discussed unidirectional CH-Potentials.
In this journal version, we also discuss a bidirectional variant and introduce an improved pruning criterion for symmetrical bidirectional A*.
The improved pruning is not specific to CH-Potentials and can also be used with other heuristics.
Compared to the conference paper, the journal version contains a significantly expanded experimental evaluation.

The rest of the article is organized as follows.
In Section~\ref{sec:related_work}, we discuss related works on goal directed search and extensions for realistic applications for hierarchical techniques.
After establishing preliminaries in Section~\ref{sec:preliminaries}, we introduce the incremental many-to-one problem and Lazy RPHAST, our algorithm to solve it efficiently, in Section~\ref{sec:lazy-rphast}.
CH-Potentials, our new A* heuristic is introduced in Section~\ref{sec:main-algo}.
There, we also discuss our improvements to the handling of low-degree nodes (Section~\ref{sec:low-deg-improvment}) and for symmetrical bidirectional A* (Section~\ref{sec:bidir_astar}).
% Section~\ref{sec:extensions} presents a multitude of applications of CH-Potentials.
In Section~\ref{sec:extensions}, we demonstrate CH-Potential's flexibility, by describing how to apply the approach to different practical applications.
Finally, in Section~\ref{sec:experiments}, we present an experimental evaluation of our approach.

\section{Related Work}\label{sec:related_work}

@@

A considerable amount of research and engineering effort has been put into studying the combination of traffic predictions with CH.
Several papers~\cite{bdsv-tdch-09,bgns-tdcha-10,klsv-dtdch-10,bgsv-mtdtt-13} and an entire dissertation~\cite{b-tdrpc-14} have been published on the subject.
Different variants with trade-offs regarding exactness, query speed and space consumption were proposed~\cite{bgsv-mtdtt-13}.
Recently, a new approach has been published~\cite{swz-sfert-21} which simultaneously achieves competitive results in all three aspects but only at the cost of considerable implementation complexity.

@@

In~\cite{dsw-cch-15}, CH is extended to Customizable CH (CCH).
CCH also has a second preprocessing phase where weights can be altered.
Supporting turn costs in CCH was studied in~\cite{bwzz-cchtc-20}.
% Turn integration still causes a slowdown of at least factor three to the second preprocessing phase.

@@

While these works show that it is possible to extend hierarchical approaches, they also show that it is non-trivial.
Further, in every extension the flexibility available at query time is fairly limited.
Combining these hierarchical extensions is an unsolved problem.

CH-Potentials is not the first work to combine hierarchical approaches and A*~\cite{bdsssw-chgds-10,gkw-blwr-07,bdgwz-sfpcs-19}.
However, previous works mostly focused on accelerating hierarchical approaches further rather than exploiting A*'s flexibility.

ALT~\cite{gh-cspas-05,gw-cppsp-05} and CPD-Heuristics~\cite{DBLP:conf/ijcai/BonoGHS19} are the two techniques with high conceptual similarity to CH-Potentials.
% Just as our approach, ALT is A*-based.
ALT has been combined with shortcuts~\cite{bdsssw-chgds-10} and also extended for dynamic graphs~\cite{dw-lbrdg-07} and time-dependent routing~\cite{ndls-bastd-12,dn-crdtd-12}.
% However, ALT's heuristic is not preprocessing-tight.
%
CPD-Heuristics are a combination of A* and Compressed Path Databases (CPD).
A CPD can quickly compute the first edge of a shortest path between any two nodes.
In~\cite{DBLP:conf/ijcai/BonoGHS19}, SRC~\cite{DBLP:conf/socs/StrasserHB14} is used as CPD.
For every distance estimation, a shortest path to the target is computed, whose length is used as the heuristic value.
Unfortunately, the employed CPD's quadratic preprocessing running time is problematic on large street networks.
%
% MtsCopa~\cite{DBLP:journals/tciaig/BaierBHH15} uses a similar idea to CPD-Heuristics but is only applied in a hunter-prey application.
%
In~\cite{DBLP:conf/ijcai/0002UJAKK18} the weighted graph is embedded into Euclidean space using FastMap. % such that distances in space and distances in the graph roughly correspond.
The Euclidean distance is then used as a distance estimate for A*.
% This results in a good but not a preprocessing-tight heuristic.

@@

We consider directed graphs $G=(V,E)$ with $n=|V|$ nodes and $m=|E|$ edges with weight functions $w : E \to \mathbb{R}$.
Given nodes $s$ and $t$ we want to obtain a $st$-path $P=(s=v_0,\dots,t=v_k)$ of minimum weight $w(P) := \sum_{i} w(v_{i-1}v_i)$.
We denote this shortest path weight as the distance $\dist_w(s,t)$ between $s$ and $t$.
If there is no path, we set $\dist_w(s,t)=\infty$.

@@

In this section, we recall the central aspects of Dijkstra's classical shortest path algorithm~\cite{d-ntpcg-59} and introduce notation used throughout this paper.
Dijkstra's algorithm maintains an array of tentative distances $D$ and a priority queue $Q$ of nodes ordered by increasing distance from $s$.
Initially, all distances are set to $\infty$.
The start node distance $D[s]$ is set to zero and the queue is initialized with $s$.
In each iteration, the closest remaining node $u$ is popped from the queue and \emph{settled}.
Outgoing edges $uv$ are relaxed, i.e. $D[v]$ is set to $\min(D[v], D[u] + w(uv))$.
If the tentative distance at the head node $D[v]$ was improved, it's added to the queue or its position in the queue is adjusted.
Once $t$ was settled, $D[t]$ contains the shortest path distance between $s$ and $t$.
The search can stop when $t$ is settled.

Dijkstra's algorithm can also be run from the target $t$ on the reversed graph.
We call this a \emph{backward search}.
Running two Dijkstra searches simultaneously, one from $s$ and a backward search from $t$, until the searches meet is called \emph{bidirectional} search.
Typically, the searches are interleaved by alternating settling a node from each direction.
Another common approach is to always settle a node from the direction with the smaller minimum queue key.
Theoretically, any direction interleaving strategy is possible.
When the sum of the minimum keys in both queues $\overrightarrow{k} + \overleftarrow{k}$ is greater than the so far best known tentative total distance $\mu$ the algorithm can terminate.

@@

A* is a goal-directed variant of Dijkstra's algorithm.
% A* uses a \emph{heuristic} to guide the search towards the goal.
It uses a heuristic function $h_t(v)$ that maps a node $v$ onto an estimate of the distance from $v$ to the target $t$.
A* orders the nodes in the priority queue by $D[v] + h_t(v)$ instead of $D[v]$ as Dijkstra's algorithm does it.
This results in nodes closer to the target being explored earlier.
Dijkstra's algorithm is a special case of A* with $h_t(v)=0$.
A* is equivalent to running Dijkstra's algorithm on the graph with \emph{reduced weights}~\cite{hnr-afbhd-68}.
This reduced weight function is defined as $w_{h_t}(uv) := w(uv) - h_t(u) + h_t(v)$.
A heuristic function is called \emph{feasible}, if $w_{h_t}(e) \geq 0$ for all edges $e$.
If the employed heuristic is feasible, A* is guaranteed to have found the optimal distance after settling $t$.
In this article, we limit our discussion to feasible heuristics.
% TODO more explicit details on A* correctness?

@@

\node (s) at (0,0)     [node] {$s$};
\node (2) at (1.5,1.5) [node] {};
\node (3) at (3.7,1)   [node] {};
\node (4) at (5,3)     [node] {$m$};
\node (5) at (5.75,-0.1)  [node] {};
\node (6) at (6.75,1.2)   [node] {};
\node (7) at (7.75,-0.5)    [node] {};
\node (8) at (8.5,2)   [node] {};
\node (t) at (10,0.5)  [node] {$t$};

@@

\draw[dashed] (4) -- (6);
\draw[dashed] (6) -- (8);
\end{tikzpicture}
\caption{
Solid lines are edges in $G$. Dotted lines are shortcuts. Red is shortest $st$-path in $G$. Blue is equaly long up-down $st$-path in $G^+$. $m$ is the mid node.
}
\label{fig:ch}
\Description[Nodes and edges of a shortest path in the original graph and the corresponding path in the augmented graph.]{
Nodes and edges of a shortest path in the original graph and the corresponding path in the augmented graph.
The y-position on the figure indicates the rank of each node.
Where the original shortest path down and then up, a shortcut edge connects the two higher nodes.
The path in the augmented graph goes only up and then down.
}
\end{figure}

A CH is a two-phase technique to efficiently compute exact, shortest paths.
For details, we refer to \cite{gssv-erlrn-12,dsw-cch-15}.
In this section, we give an introduction.

\begin{algorithm2e}
\KwData{$B[v]$: tentative distance from $v$ to target $t$}
\KwData{Minimum priority queue $Q$, also called open list, ordered by tentative distances}
$B[v] \gets +\infty$ for all $v\neq t$;
$B[t] \gets 0$\;
Make $Q$ only contain $t$ with weight $0$\;
\While{$Q$ not empty}{
  $u\gets$ pop minimum element from $Q$\;
  \For{all down-edges $vu$ in $G^+_\ell$}{
    \If{$B[v] > w_\ell(vu) + B[u]$}{
      $B[v]\gets w_\ell(vu) + B[u]$\;
                        Add $v$ or decrease $v$'s key in $Q$ to $B[v]$\;
    }
  }
}
\caption{CH backward search}
\label{algo:ch-backward}
\end{algorithm2e}

In a CH, each node gets a unique rank representing its ``importance''.
By convention, important nodes are usually depicted at the ``top'' while unimportant nodes are positioned at the ``bottom''.
The intuition is that dead-ends are unimportant while highways or bridges are important.
An edge goes \emph{up} when its tail node has lower rank than its head node.
\emph{Down} edges are defined analogously.
An \emph{up-down path} is a path where only one node $m$ is more important than both its neighbors.
$m$ is called the \emph{mid} node.
An \emph{up path} is a path where the last node is the mid node.
Similarly, for a \emph{down path}, the first node has to be the mid node.
% Every up and down path is an up-down path.
%
In the preprocessing phase, first, a node ordering is computed heuristically.
Then, \emph{shortcut} edges are added to the input graph $G$ to obtain $G^+$.
This is done by repeatedly contracting the least important node and adding shortcuts between its higher ranked neighbors.
%See~\cite{gssv-erlrn-12} for the details.
After the preprocessing, for every pair of nodes $s$ and $t$ there exists a shortest up-down $st$-path in $G^+$ with the same length as a shortest path in $G$.
See Figure~\ref{fig:ch} for a proof sketch.
From every shortest path (red) in $G$, an up-down path of equal length in $G^+$ (blue) can be constructed.
Thus, we can restrict our search to up-down paths in $G^+$.
The search is bidirectional.
The forward search starts from $s$ and only follows up-edges.
Similarly, the backward search starts at $t$ and only follows down-edges in reversed direction.
The two searches meet at the mid node.
Pseudocode for the backward search, i.e., the path from $m$ to $t$, is presented in Algorithm~\ref{algo:ch-backward}.
The forward search works analogously.
%
A CH query is fast, if the number of nodes reachable via only up- or down-nodes is small.
On road networks, this is the case~\cite{gssv-erlrn-12,dgpw-crprn-13}.
On graphs with low treewidth, this is also the case~\cite{dsw-cch-15,hs-gbpo-18}.

>>>>

\newcommand*{\dist}{\operatorname{dist}}
\newcommand*{\gchu}{G^{\uparrow}}
\newcommand*{\gchd}{G^{\downarrow}}
\newcommand*{\rgchd}{\overleftarrow{G^{\downarrow}}}
\newcommand*{\rgchu}{\overleftarrow{G^{\uparrow}}}
\newcommand*{\echu}{E^{\uparrow}}
\newcommand*{\echd}{E^{\downarrow}}
\newcommand*{\rechd}{\overleftarrow{E^{\downarrow}}}
\newcommand*{\rechu}{\overleftarrow{E^{\uparrow}}}

@@

\author{Ben Strasser}
\email{academia@ben-strasser.net}
\affiliation{%
  \institution{Institute of Theoretical Informatics, Algorithmics I, Karlsruhe Institute of Technology}
  \city{Karlsruhe}
  \country{Germany}
}

@@

\begin{abstract}
We study exact, efficient and practical algorithms for route planning applications in large road networks.
Routing applications often require integrating the current traffic situation, planning ahead with traffic predictions for the future, respecting forbidden turns, and many other features depending on the specific application.
While Dijkstra's algorithm can be used to solve these problems, it is too slow for many applications.
A* is a classical approach to accelerate Dijkstra's algorithm.
A* can support many extended scenarios without much additional implementation complexity.
However, A*'s performance depends on the availability of a good heuristic that estimates distances.
Computing tight distance estimates is a challenge on its own.
On road networks, shortest paths can also be quickly computed using hierarchical speedup techniques.
They achieve speed and exactness but sacrifice A*'s flexibility.
Extending them to certain practical applications can be hard.
In this paper, we present Lazy RPHAST, an algorithm to efficiently compute distances from many vertices towards a single target and use it as an A* heuristic.
As it uses Contraction Hierarchies (CH), a hierarchical technique, we call our A* heuristic CH-Potentials.
Our approach allows decoupling the supported applications from the hierarchical speed-up technique.
Additionally, we describe A* optimizations to accelerate the processing of low degree vertices, which often occur in road networks and present a new pruning criterion for symmetrical bidirectional A*.
An extensive experimental study confirms the practical applicability of our approach for many applications.
\end{abstract}

@@

\section{Introduction}
\label{sec:intro}
The past decade has seen a plethora of research works on route planning in large road networks~\cite{bdgmpsww-rptn-16}.
Routing a user through a road network can be formalized as the shortest path problem in weighted graphs.
Vertices represent intersections.
Roads are modeled using edges.
Edges are weighted by their traversal times.
The problem can be solved with Dijkstra's algorithm~\cite{d-ntpcg-59}.
Unfortunately, on continental sized networks, it is too slow for many applications.
Thus, speed-up techniques have been developed.
One popular example are Contraction Hierarchies~(CH)~\cite{gssv-erlrn-12}.
They have been used successfully in many real world applications.
A CH exploits the inherent hierarchy of road networks.
In a preprocessing step, additional shortcut edges are inserted, which allow skipping unimportant parts of the network at query time.
Another popular example is Multi-Level-Dijkstra~(MLD)~\cite{swz-umlgt-02} also known as CRP~\cite{dgpw-crprn-13}.
It is also used in practice~\cite{bingblog}.
MLD also uses shortcut edges.
Both approaches achieve speed-ups of at least three orders of magnitude over Dijkstra's algorithm.

\begin{figure}
\centering
\includegraphics[width=.6\columnwidth]{fig/searchspace_st.png}
\caption{Vertices explored by A*. Color indicates the vertex removal order from the queue. Blue was removed first. Next is green. Red was removed last.}
\label{img:search-space}
\Description[A map where vertices visited by A* are marked.]{
A map of Karlsruhe where vertices visited by A* are marked.
The shape of the area with the visited vertices is similar to a drop.
The source vertex roughly lies in the center of the drop, the target vertex at the tip.
}
\end{figure}

Unfortunately, for many real world applications, this basic graph model is too simplistic.
For realistic routing, many additional features need to be considered.
This includes turn costs and restrictions, live traffic, user preferences, and traffic predictions. %, temporary driving bans for certain types of vehicles.
Some applications may have additional application-specific requirements.
Extending Dijkstra's algorithm to support these features is usually easy.
Extending hierarchical speed-up techniques is often possible, too.
However, the algorithm development is vastly more complex. % TODO cite some papers
For every feature, dedicated research papers exist that extend CH and other approaches.
Supporting the combination of several features is even harder.
For example, we are not aware of any work combining all features mentioned above.
In this article, we describe an algorithmic building block, that allows handling the combination of all above-mentioned features.

Our approach decouples the development of extensions from the hierarchical speed-up technique by utilizing the A* algorithm~\cite{hnr-afbhd-68}.
A* is a goal-directed variant of Dijkstra's algorithm.
See Figure~\ref{img:search-space} for an example of vertices traversed during an A* search.
A* uses a \emph{heuristic} to guide the search towards the goal.
A heuristic is a function that maps a vertex to an estimate of the remaining distance to the goal.
A*'s running time crucially depends on how tight this estimate is.
Further, evaluating the heuristic must be fast.
% The free flow travel time to the goal is a valid heuristic for most features.
% In many settings, it is the tightest lower bound available during preprocessing, i.e., independent of real-time data and user settings.
% However, to be useful we must be able to evaluate the free flow heuristic \todo{ist free flow heuristic besser als tight heuristic?} quickly and exactly.
In this article, we introduce Lazy RPHAST and build on it CH-Potentials, a fast heuristic with tight estimates.
Internally, the heuristic uses a CH.
Fortunately, this is an implementation detail from the perspective of the A*.
To support a new feature, we only need to modify the A* algorithm.
The heuristic core containing the CH remains untouched.
Extending A* is vastly easier than extending a CH.
This enables us to design algorithms for a multitude of features.

% TODO rewrite, make self contained
The paper introduced the basic CH-Potentials heuristic, some query optimizations for handling of low-degree vertices, common in road networks, and some applications of CH-Potentials.
In this article, we discuss and evaluate Lazy RPHAST, the central ingredient of CH-Potentials, in greater detail and introduce an improved variant for Customizable Contraction Hierarchies.
% Furthermore, we evaluate Lazy RPHAST in the context of one-to-many and many-to-many problems.
The original paper only discussed unidirectional CH-Potentials.
In this journal version, we also discuss a bidirectional variant and introduce an improved pruning criterion for symmetrical bidirectional A*.
The improved pruning is not specific to CH-Potentials and can also be used with other heuristics.
Compared to the conference paper, the journal version contains a significantly expanded experimental evaluation.

The rest of the article is organized as follows.
In Section~\ref{sec:related_work}, we discuss related works on goal directed search and extensions for realistic applications for hierarchical techniques.
After establishing preliminaries in Section~\ref{sec:preliminaries}, we introduce the incremental many-to-one problem and Lazy RPHAST, our algorithm to solve it efficiently, in Section~\ref{sec:lazy-rphast}.
CH-Potentials, our new A* heuristic is introduced in Section~\ref{sec:main-algo}.
There, we also discuss our improvements to the handling of low-degree vertices (Section~\ref{sec:low-deg-improvment}) and for symmetrical bidirectional A* (Section~\ref{sec:bidir_astar}).
% Section~\ref{sec:extensions} presents a multitude of applications of CH-Potentials.
In Section~\ref{sec:extensions}, we demonstrate flexibility of our approach, by describing how to apply the approach to different practical applications.
Finally, in Section~\ref{sec:experiments}, we present an experimental evaluation of our approach.

\subsection{Related Work}\label{sec:related_work}

@@

A considerable amount of research and engineering effort has been put into studying the combination of traffic predictions with CH.
Several papers~\cite{bdsv-tdch-09,bgns-tdcha-10,klsv-dtdch-10,bgsv-mtdtt-13} and an entire dissertation~\cite{b-tdrpc-14} have been published on the subject.
Different variants with trade-offs regarding exactness, query speed and space consumption were proposed~\cite{bgsv-mtdtt-13}.
Recently, a new approach has been published~\cite{swz-sfert-21} which simultaneously achieves competitive results in all three aspects but only at the cost of considerable implementation complexity. % TODO argue what is complicated to implement?

@@

In~\cite{dsw-cch-15}, CH is extended to Customizable CH (CCH). % TODO paragraph looks ugly, somehow make more compact
CCH also has a second preprocessing phase where weights can be altered.
Supporting turn costs in CCH was studied in~\cite{bwzz-cchtc-20}.
Turn integration still causes a slowdown of at least factor three to the second preprocessing phase.

@@

Clearly, extending hierarchical approaches is possible, but it requires significant research effort.
% While these works show that it is possible to extend hierarchical approaches, they also show that it is non-trivial. % TODO Reviewer1 similarly, discuss why "they also show it is non trivial". What does this mean?
Further, extensions to these speedup techniques cannot easily be composed with each other.
Combining these extensions is an unsolved problem.

CH-Potentials is not the first technique to combine hierarchical approaches and A*~\cite{bdsssw-chgds-10,gkw-blwr-07,bdgwz-sfpcs-19}.
However, previous works mostly focused on accelerating hierarchical approaches further rather than exploiting A*'s flexibility.

ALT~\cite{gh-cspas-05,gw-cppsp-05} and CPD-Heuristics~\cite{DBLP:conf/ijcai/BonoGHS19} are the two techniques with high conceptual similarity to CH-Potentials.
% Just as our approach, ALT is A*-based.
ALT has been combined with shortcuts~\cite{bdsssw-chgds-10} and also extended for dynamic graphs~\cite{dw-lbrdg-07} and time-dependent routing~\cite{ndls-bastd-12,dn-crdtd-12}.
% However, ALT's heuristic is not preprocessing-tight.
%
CPD-Heuristics are a combination of A* and Compressed Path Databases (CPD).
A CPD can quickly compute the first edge of a shortest path between any two vertices.
In~\cite{DBLP:conf/ijcai/BonoGHS19}, SRC~\cite{DBLP:conf/socs/StrasserHB14} is used as CPD.
For every distance estimation, a shortest path to the target is computed, whose length is used as the heuristic value.
Unfortunately, the employed CPD's quadratic preprocessing running time is problematic on large road networks.
%
% MtsCopa~\cite{DBLP:journals/tciaig/BaierBHH15} uses a similar idea to CPD-Heuristics but is only applied in a hunter-prey application.
%
In~\cite{DBLP:conf/ijcai/0002UJAKK18} the weighted graph is embedded into Euclidean space using FastMap. % such that distances in space and distances in the graph roughly correspond.
The Euclidean distance is then used as a distance estimate for A*.
% This results in a good but not a preprocessing-tight heuristic.

\subsection{Contribution and Outline}

In this work, we introduce Lazy RPHAST, a simple and efficient CH-based algorithm for the incremental many-to-one shortest path problem (Section~\ref{sec:lazy-rphast}).
In Section~\ref{sec:main-algo}, we present CH-Potentials, a novel fast and tight heuristic for A* in road networks build on Lazy RPHAST.
Further, we present several useful optimizations for A* in road networks accelerating the processing of low degree vertices (Section~\ref{sec:low-deg-improvment}) and an improved variant of bidirectional A* (Section~\ref{sec:bidir_astar}).
In Section~\ref{TODO}, we introduce a unified framework to apply CH-Potentials to a variety of practical route planning problems without any modifications to the preprocessing.
Finally, an extensive experimental evaluation in Section~\ref{sec:experiments} confirms the applicability of our results.

@@

We consider directed graphs $G=(V,E)$ with $E \subseteq V \times V$ with $n=|V|$ vertices and $m=|E|$ edges with weight functions $w : E \to \mathbb{R}^{\geq 0}$.
We use $uv$ as a short notation for the edge $(u,v)$.
The \emph{reversed} graph $\overleftarrow{G} := (V, \{ vu \mid uv \in E \})$ contains all edges in reverse direction.
The corresponding reversed weight function is $\overleftarrow{w}(vu) := w(uv)$.
We denote by $N(u) = \{ v \mid uv \in E \vee vu \in E \}$ the \emph{undirected neighborhood} of a vertex $u$ and refer to the numbers of distinct neighbors $|N(u)|$ as the \emph{degree} of $u$.

Given vertices $s$ and $t$ we want to obtain a $st$-path $P=(s=v_0,\dots,t=v_k)$ of minimum weight $w(P) := \sum_{i} w(v_{i-1}v_i)$.
We denote this shortest path weight as the distance $\dist_w(s,t)$ between $s$ and $t$.
If there is no path, we set $\dist_w(s,t)=\infty$. % TODO edge notation, degree (undirected)

@@

In this section, we recall the central aspects of Dijkstra's classical shortest path algorithm~\cite{d-ntpcg-59} and introduce notation used throughout this paper.
Dijkstra's algorithm maintains an array of tentative distances $D$ and a priority queue $Q$ of vertices ordered by increasing distance from $s$.
We denote by $k$ the minimum key in $Q$, i.e.\ the tentative distance of the closest remaining vertex.
Initially, all distances are set to $\infty$.
The start vertex distance $D[s]$ is set to zero and the queue is initialized with $s$.
In each iteration, the closest remaining vertex $u$ is popped from the queue and \emph{settled}.
Outgoing edges $uv$ are relaxed, i.e. $D[v]$ is set to $\min(D[v], D[u] + w(uv))$.
If the tentative distance at the head vertex $D[v]$ was improved, it's added to the queue or its position in the queue is adjusted.
Once $t$ was settled, $D[t]$ contains the shortest path distance between $s$ and $t$.
The search can stop when $t$ is settled.
The vertices visited by Dijkstra's algorithm throughout the search are denoted as the \emph{search space}.

Dijkstra's algorithm can also be run from the target $t$ on the reversed graph $\overleftarrow{G}$. % TODO introduce reverse dist etc
We call this a \emph{backward search}.
Running two Dijkstra searches simultaneously, one from $s$ and a backward search from $t$, until the searches meet is called \emph{bidirectional} search.
In this case, we denote by $\overrightarrow{D}$, $\overrightarrow{Q}$ and $\overrightarrow{k}$ the distances, queue and minimum queue key of the forward search and by $\overleftarrow{D}$, $\overleftarrow{Q}$ and $\overleftarrow{k}$ the respective data of the backward search.
Typically, the searches are interleaved by alternating settling a vertex from each direction.
Another common approach is to always settle a vertex from the direction with the smaller minimum queue key.
Theoretically, any direction interleaving strategy is possible.
When the sum of the minimum keys in both queues $\overrightarrow{k} + \overleftarrow{k}$ is greater than the so far best known tentative total distance $\mu$ the algorithm can terminate.

@@

A* is a goal-directed variant of Dijkstra's algorithm.
% A* uses a \emph{heuristic} to guide the search towards the goal.
It uses a heuristic function $h_t(v)$ that maps a vertex $v$ onto an estimate of the distance from $v$ to the target $t$.
A* orders the vertices in the priority queue by $D[v] + h_t(v)$ instead of $D[v]$ as Dijkstra's algorithm does it.
This results in vertices closer to the target being explored earlier.
Dijkstra's algorithm is a special case of A* with $h_t(v)=0$.
A* is equivalent to running Dijkstra's algorithm on the graph with \emph{reduced weights}~\cite{hnr-afbhd-68}.
This reduced weight function is defined as $w_{h_t}(uv) := w(uv) - h_t(u) + h_t(v)$.
A heuristic function is called \emph{feasible}, if $w_{h_t}(e) \geq 0$ for all edges $e$.
If the employed heuristic is feasible, A* is guaranteed to have found the optimal distance after settling $t$.
In this article, we limit our discussion to feasible heuristics.
% TODO more explicit details on A* correctness?

@@

\node (s) at (0,0)     [node] {$s$};
\node (2) at (1.5,1.5) [node] {};
\node (3) at (3.7,1)   [node] {};
\node (4) at (5,3)     [node] {$v^*$};
\node (5) at (5.75,-0.1)  [node] {};
\node (6) at (6.75,1.2)   [node] {};
\node (7) at (7.75,-0.5)    [node] {};
\node (8) at (8.5,2)   [node] {};
\node (t) at (10,0.5)  [node] {$t$};

@@

\draw[dashed] (4) -- (6);
\draw[dashed] (6) -- (8);
\end{tikzpicture}
\caption{
Solid lines are edges in $G$. Dotted lines are shortcuts. Red is a shortest $st$-path in $G$. Blue is equally long up-down $st$-path in $G^+$. $v^*$ is the mid vertex. % TODO different letter for m
}
\label{fig:ch}
\Description[Vertices and edges of a shortest path in the original graph and the corresponding path in the augmented graph.]{
Vertices and edges of a shortest path in the original graph and the corresponding path in the augmented graph.
The y-position on the figure indicates the rank of each vertex.
Where the original shortest path down and then up, a shortcut edge connects the two higher vertices.
The path in the augmented graph goes only up and then down.
}
\end{figure}

A CH is a two-phase technique to efficiently compute exact, shortest paths on road networks.
For details, we refer to \cite{gssv-erlrn-12,dsw-cch-15}.
In this section, we give an introduction.

\begin{algorithm2e}[tb]
\KwData{$D^{\downarrow}[v]$: tentative distance from any vertex $v \in V$ to target $t$ in $\gchd$}
\KwData{Minimum priority queue $Q$, ordered by tentative distances}
$D^{\downarrow}[v] \gets +\infty$ for all $v\neq t$\;
$D^{\downarrow}[t] \gets 0$\;
Make $Q$ only contain $t$ with tentative distance $0$\;
\While{$Q$ not empty}{
  $u\gets$ pop minimum element from $Q$\;
  \For{all down-edges $vu \in \echd$}{
    \If{$D^{\downarrow}[v] > w(vu) + D^{\downarrow}[u]$}{
      $D^{\downarrow}[v]\gets w(vu) + D^{\downarrow}[u]$\;
      Add $v$ or decrease $v$'s key in $Q$ to $D^{\downarrow}[v]$\;
    }
  }
}
\caption{CH backward search.}
\label{algo:ch-backward}
\end{algorithm2e}

In a CH, each vertex gets a unique rank representing its ``importance''.
By convention, important vertices are usually depicted at the ``top'' while unimportant vertices are positioned at the ``bottom''.
The intuition is that dead-ends are unimportant while highways or bridges are important.
An edge goes \emph{up} when its tail vertex has lower rank than its head vertex.
\emph{Down} edges are defined analogously.
An \emph{up-down path} is a path where only one vertex $v^*$ is more important than both its neighbors.
$v^*$ is called the \emph{mid} vertex.
An \emph{up path} is a path where the last vertex is the mid vertex.
Similarly, for a \emph{down path}, the first vertex has to be the mid vertex.
% Every up and down path is an up-down path.
%
In the preprocessing phase, first, a vertex ordering is computed heuristically.
Then, \emph{shortcut} edges are added to the input graph $G$ to obtain $G^+$.
This is done by repeatedly contracting the least important vertex and adding shortcuts between its higher ranked neighbors.
%See~\cite{gssv-erlrn-12} for the details.
After the preprocessing, for every pair of vertices $s$ and $t$ there exists a shortest up-down $st$-path in $G^+$ with the same length as a shortest path in $G$.
See Figure~\ref{fig:ch} for an illustration.
From every shortest path (red) in $G$, an up-down path of equal length in $G^+$ (blue) can be constructed.
Thus, we can restrict our search to up-down paths in $G^+$.
The search is bidirectional.
The forward search starts from $s$ and only follows up-edges.
Similarly, the backward search starts at $t$ and only follows down-edges in reversed direction.
By construction, the mid vertex must be in the intersection of the forward and backward search spaces.
Pseudocode for the backward search is presented in Algorithm~\ref{algo:ch-backward}.
The forward search works analogously.
% TODO \gchu and \gchd
